Geospatial - Assignment 3

### 2025-02-14 hw3/hw3_MoritzPeistNoemiLucchiSimonVellin.Rmd
Small formatting for ex2 part 1 and corrected file paths

## Data Wrangling
From the R package spData, we use the world dataset to extract the geometry of Spain, filtering only for continental Spain. Then, from the dataset containing the most populated cities, we extract only the cities in Spain, sort them in decreasing order of population, and retain only the top 10. Finally, from these top 10 cities, we extract Madrid and Vigo, which will be used in the plot.

```{r}
# Filter for continental Spain only 
spain <- world %>% 
  filter(name_long=='Spain') %>%
  st_cast("POLYGON") %>%                   
  slice_max(st_area(.))

@@

# Filter Madrid and Vigo for the plot 
madrid_vigo <- spain_top_10_cities %>%
  filter(NAME %in% c("Madrid", "Vigo"))

```

## Methodological Approaches
First, we transform the CRS from EPSG:4326 to EPSG:3035, as the former uses latitude and longitude, which would lead to incorrect distance calculations in kilometers. Next, we intersect the road network with Spain’s boundaries to restrict the analysis to roads that are actually within Spain. We then create an empty raster over Spain, set a high resolution, and define the CRS as EPSG:3035 for the same reason mentioned earlier. After that, we generate the friction surface by assigning a value of 1 to cells that contain roads and 100 to cells that do not. This ensures that areas without roads have higher friction, influencing the computation of the shortest path between cities. Subsequently, we compute the transition matrix and, finally, calculate and print the distance matrix, which includes all pairwise distances between the top 10 most populated Spanish cities.

@@

# Create transition matrix for distance calculation
tr_matrix <- transition(raster(road_raster), 
                        transitionFunction = mean, 
                        directions = 8)

@@

for(i in 1:n_cities) {
  for(j in 1:n_cities) {
    if(i < j) {  
      path <- shortestPath(tr_matrix, 
                           origin = as_Spatial(cities_planar[i,]), 
                           goal = as_Spatial(cities_planar[j,]), 
                           output = "SpatialLines")
      path_sf <- st_as_sf(path)
      st_crs(path_sf) <- st_crs(cities_planar)
      # Calculate distance in kilometers
      dist <- as.numeric(st_length(path_sf)) / 1000
      dist_matrix[i,j] <- dist
      dist_matrix[j,i] <- dist}}}

@@

## Plotting 
After these manipulations, we can plot continental Spain with its roads, highlighting Madrid and Vigo, as shown in the reference plot.

```{r}
# Plotting
ggplot() +
  geom_sf(data = st_transform(spain, 4326), fill = "lightgray") +
  geom_sf(data = st_transform(spain_roads, 4326), color = "black", size = 0.3) +
  geom_sf(data = st_transform(spain_top_10_cities, 4326), size = 1, color = "black") +
  geom_sf(data = st_transform(madrid_vigo %>% filter(NAME == "Madrid"), 4326), 
          aes(shape = "Madrid", color = "Madrid"), size = 3) +
  geom_sf(data = st_transform(madrid_vigo %>% filter(NAME == "Vigo"), 4326), 
          aes(shape = "Vigo", color = "Vigo"), size = 3) +
  scale_color_manual(name = "Cities",
                     values = c("Madrid" = "red", "Vigo" = "turquoise")) +
  scale_shape_manual(name = "Cities",
                     values = c("Madrid" = 16, "Vigo" = 17)) 
```

>>>>

From the R package spData, we use the world dataset to extract the geometry of Spain, filtering only for continental Spain. Then, from the dataset containing the most populated cities, we extract only the cities in Spain, sort them in decreasing order of population, and retain only the top 10. Finally, from these top 10 cities, we extract Madrid and Vigo, which will be used in the plot.

```{r}
# Filter for continental Spain only
spain <- world %>%
  filter(name_long == "Spain") %>%
  st_cast("POLYGON") %>%
  slice_max(st_area(.))

@@

# Filter Madrid and Vigo for the plot
madrid_vigo <- spain_top_10_cities %>%
  filter(NAME %in% c("Madrid", "Vigo"))
```

## Methodological Approach

First, we transform the CRS from EPSG:4326 to EPSG:3035, as the former uses latitude and longitude, which would lead to incorrect distance calculations in kilometers. Next, we intersect the road network with Spain’s boundaries to restrict the analysis to roads that are actually within Spain. We then create an empty raster over Spain, set a high resolution, and define the CRS as EPSG:3035 for the same reason mentioned earlier. After that, we generate the friction surface by assigning a value of 1 to cells that contain roads and 100 to cells that do not. This ensures that areas without roads have higher friction, influencing the computation of the shortest path between cities. Subsequently, we compute the transition matrix and, finally, calculate and print the distance matrix, which includes all pairwise distances between the top 10 most populated Spanish cities.

@@

# Create transition matrix for distance calculation
tr_matrix <- transition(raster(road_raster),
  transitionFunction = mean,
  directions = 8
)

@@

for (i in 1:n_cities) {
  for (j in 1:n_cities) {
    if (i < j) {
      path <- shortestPath(tr_matrix,
        origin = as_Spatial(cities_planar[i, ]),
        goal = as_Spatial(cities_planar[j, ]),
        output = "SpatialLines"
      )
      path_sf <- st_as_sf(path)
      st_crs(path_sf) <- st_crs(cities_planar)
      # Calculate distance in kilometers
      dist <- as.numeric(st_length(path_sf)) / 1000
      dist_matrix[i, j] <- dist
      dist_matrix[j, i] <- dist
    }
  }
}

@@

## Plotting

After these manipulations, we can plot continental Spain with its roads, highlighting Madrid and Vigo, as shown in the reference plot.

```{r}
# Plotting
ggplot() +
  geom_sf(data = st_transform(spain, 4326), fill = "lightgray") +
  geom_sf(data = st_transform(spain_roads, 4326), color = "black", size = 0.3) +
  geom_sf(data = st_transform(spain_top_10_cities, 4326), size = 1, color = "black") +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Madrid"), 4326),
    aes(shape = "Madrid", color = "Madrid"), size = 3
  ) +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Vigo"), 4326),
    aes(shape = "Vigo", color = "Vigo"), size = 3
  ) +
  scale_color_manual(
    name = "Cities",
    values = c("Madrid" = "red", "Vigo" = "turquoise")
  ) +
  scale_shape_manual(
    name = "Cities",
    values = c("Madrid" = 16, "Vigo" = 17)
  )
```
